Dissertation Proposal

### 2023-10-26 proposal.tex
complete
clean up

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}

@@

  In this paper, we present a curated list of misconceptions, and an
  instrument to detect them. These are distilled from student work
  over several years and match and extend prior research.  We also
  present an automated, self-guided tutoring system. The tutor builds
  on strategies in the education literature and is explicitly designed
  around identifying and correcting misconceptions.

  We have tested the tutor in multiple settings.
  Our data consistently show that (a) the misconceptions we tackle are
  widespread, and (b) the tutor appears to improve understanding.
\end{abstract}

@@

\begin{description}
  \item[RQ1] What are the misconceptions that are about program
  behavior, apply to most modern programming languages, and present
  (even) in students with prior programming backgrounds?
  \item[RQ2] What would be a tutoring system (the SMoL Tutor) that
  draws on cognitive and educational psychology concepts and that aim
  at correcting the misconceptions?
  \item[RQ3] How effective is SMoL Tutor at correcting the
  misconceptions?
\end{description}

@@

  \item We provide a collection of brief programs for detecting
        misconceptions about the features of SMoL and their
        interaction. These programs are short and most are readily
        translatable to a wide variety of programming languages that
        share the SMoL characteristics.

  \item We show that multiple populations have problems with these
        programs.

  \item We present a list of curated misconceptions generated
        after multiple rounds of analysis.

  \item We implement these misconceptions as interpreters, which
        find weaknesses in our manual analysis and hold potential for
        generative use in the future.

  \item We present a tutoring system, the SMoL Tutor, that measurably
        corrects these misconceptions.

@@

In the near future, I would like to

\paragraph{A Note on Terminology} We use the term
``behavior'' to refer to the meaning of programs in terms of the
answers they produce. A more standard term for this would, of course, be
``semantics''. However, the term ``semantics tutor'' might
mislead some readers into thinking it
teaches people to read or write a formal semantics, e.g., an
introduction to ``Greek'' notation. Because that is not the kind of
tutor we are describing, to avoid confusion, we use the term
``behavior'' instead.

@@

There is an extensive body of literature on tutoring systems (\cite{vanlehnBehaviorTutoringSystems2006a} is
a quality survey), and indeed whole conferences are dedicated to
them. We draw on this literature. In particular, it is common in the
literature to talk about a ``two-loop'' architecture \cite{vanlehnBehaviorTutoringSystems2006a} where the outer
loop iterates through ``tasks'' (i.e., educational activities) and the inner loop iterates through UI events within a task. We follow the same
structure in our tutor (\Cref{s:tutor-uiux}).

Many tutoring systems focus on teaching programming
(such as the well-known and heavily studied LISP Tutor \cite{andersonLISPTutor1985}), and in
the process undoubtedly address some program behavior
misconceptions. Our SMoL Tutor differs in a notable way: it does not
try to teach programming per se. Instead, it assumes a basic
programming background and focuses entirely on program behavior misconceptions
and correcting them. We are not aware of a tutoring system (in computer science) that has
this specific design.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:univ-1-course}) using Racket
\cite{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
We have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

Nevertheless, most SMoL programs are easy to translate to other
languages. We present machine-translated Python and JavaScript
versions of our programs in \supplRef{Translated\_Programs.html}. Furthermore, we intend to
make a multi-lingual tutor in the future (\Cref{s:future-work}).

@@

Students are given a working implementation of SMoL, built
as a \verb|#lang| language inside Racket \cite{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online
[URL anonymized].

\section{Paper Roadmap}
\label{s:roadmap}
\label{s:univ-1-course}

This paper describes a four-year effort to obtain a quality instrument
to measure misconceptions and to build a tutor to address them.

@@

We then present the tutor (\Cref{s:system-tutor}). In particular, we
present the list of misconceptions---grounded in student data---that
are the focus of this project (\Cref{s:misconceptions-tutor}). We also
evaluate the tutor's effectiveness at correcting them
(\Cref{s:effect-tutor}).

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \cite{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \cite{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \verb|sqrt(4)|?'', then \verb|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \verb|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. We expect our
  instruments to be useful in the same way.}

All these, however, add up to a somewhat challenging demand. We want
to produce a list of questions (each one an MCQ) such that
\begin{enumerate}

@@

\end{itemize}
and so on.
We therefore manually curated the Quizius output
to address these issues.

@@

We therefore turned the programs into a set of quizzes (in the US
sense: namely, a brief test of knowledge) that we call the SMoL
Quizzes. There were three quizzes, ordered by linguistic
complexity. The first consisted of only basic operators and
first-order functions. The second added variable and structure
mutation. The third added \lstinline|lambda| and higher-order functions.

@@

Question orders were partially randomized. We wanted students to get
some easy, warm-up questions initially, so those were kept at the
beginning. Similarly, we wanted programs that are syntactically
similar to stay close to each other in the quiz. This is so that, when
students got a second such program, they would not have to look far to
find the first one and confirm that they are indeed (slightly)
different, rather than wonder if they were seeing the same program
again.

@@

Due to space limitations, we present the entire instrument in \supplRef{SMoL Quizzes/instrument}.
Here we focus on a few programs where student choices correspond to
misconceptions identified earlier, thereby also showing
that the curated programs are
still effective.
Syntactically, \verb|#t| and \verb|#f| are true and false,
while \verb|#(...)| is a vector. We use a * to indicate the correct
answer (which also matches the implementation's output).

@@

So far, we have focused on identifying problems. As noted earlier
(\Cref{s:smol-quizzes}), we provided students with refutation texts after the quizzes,
but it is unclear to what extent students read, understood, or
internalized these. We also wanted to determine whether other
populations run into these issues, and it was unclear whether we would
be able to administer passive quizzes to them.

Furthermore, while the quizzes may be a useful diagnostic, our goal is
not only to find faults but to improve the understanding of basic
programming language behavior. We believe (and presumably so does
anyone else who writes a formal semantics!)\ that an understanding of
these basic program behaviors is important, and even more so when it
comes to understanding concurrency, ownership, and other advanced
features. Even more simply, misunderstanding these clearly impacts
development and debugging time.

In response, we decided to create a \emph{tutor}: the SMoL Tutor. It is built
around our quiz instruments and the detected misconceptions. We
describe the Tutor from a user's perspective in \Cref{s:tutor-uiux}, and
explain how it was populated from the
SMoL Quizzes in \Cref{s:tutor-refine-program-set}.
We then discuss what
we learned from its data (\Cref{s:misconceptions-tutor}), and finally
evaluate its effectiveness as a \emph{tutor} (\Cref{s:effect-tutor}).

@@

In addition to asking students questions, the Tutor along the way
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. We therefore refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial.
Readers can find the full Tutor in \supplRef{SMoL Tutor/instrument}.

@@

The SMoL Tutor's interpreting questions are the final instrument of
this paper.
We bootstrap it from the SMoL
Quizzes, but made the following alterations:
\begin{enumerate}

@@

  \item We renamed vector operators to avoid confusion between
        \lstinline|vset!|, which replaces a vector element, and \lstinline|set!|,
        which mutates variables. We rename all vector operators from, for example,
        \lstinline|vset!| to \lstinline|vec-set!|.

  \item We deferred the local binding forms to the end, since we felt
        they were less essential. We therefore rewrote programs that use
        it to not depend on it.

        % \item We removed some questions that we deemed ``Lispy'' (e.g.,
        %   involving the difference between parenthetical and infix syntax).

  \item We removed some programs that relied on underspecified aspects or
        unimportant aspects of SMoL. For example, some depended on
        whether operations on pairs could be applied to arbitrary vectors:

@@

  \item We resolved ambiguities in some programs, either adding
    answer choices or even adding other questions to tease
    out different interpretations.

@@

  \item We removed questions related to function equality, which was
        not a focus of the Tutor.\footnote{However, we believe there is
          a good deal of confusion about notions of equality. We
          intend to
          add that as a major tutorial component in future versions.}

  \item We removed programs of a ``Lispy'' nature, such as one where
    the answer depended on whether the reader correctly understood
    this inequality:

@@

It was important to add wrong answers that are not utterly
implausible, because those would become easy to eliminate. Therefore,
we added numeric constants mentioned in the problem, permuted some of
the values in case of multiple outputs, and so on. In general, we
ended up increasing the number of choices substantially:
only 8 out of the 89 questions have three or four choices; 70
questions have 5--8 choices; and 11 questions have 9--14 choices. We hope
this reduced both guessing and elimination, and forced students to
actually think through the program. Of course, these new answers do
not have a clear associated misconception, so their mistakes are given
the generic explanation.

@@

We now examine what we learned from the interpreting
questions in terms of program behavior
misconceptions. But first, we explain how we associate incorrect
answers with misconceptions.

We do not present the results of \textbf{begin} and
\textbf{local} (\Cref{t:tutor-goal-sentences}), because they focus on constructs not usually found in
non-Lispy languages. They were included in the Tutor to help students
write programs for the course, but are not interesting from a SMoL
perspective.

@@

By running programs through the misinterpreters, we can more uniformly
and rigorously identify \emph{all} the misconceptions associated with
a wrong answer. Furthermore, if we identify a new misconception (or
alter a misinterpreter), it is easy to automatically re-classify all
the answers. By using misinterpreters, we indeed found new
interpretations for existing program-answer pairs.
We provide all the misinterpreters in the artifact.

@@

We iteratively created
our final catalog of misconceptions
(from the perspective of the data in this paper).
We started
with misinterpreters representing the misconceptions described in
\Cref{s:misconceptions-quiz}. These are misconceptions for which we
have reasonable validation (due to the prose in Quizius), so we call
them \emph{grounded} misconceptions. We then looked at wrong answers
not covered by these but chosen by students, and did our best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we
identify with a $^\ddag$), which need to be validated in the future.\footnote{To keep
  the time spent in each module reasonable, the SMoL Tutor does not
  ask students for rationales for their programs. However, we could
  easily modify it to do so a small number of times per user, to make
  progress towards this need.} We then re-ran the misinterpreters
against the chosen answers. We terminated when all the remaining wrong
answers were either (a) found in very few students (we found a gap
between 23\% and 13\%, and hence took 14\% as the threshold),
(b) difficult for
us to attribute to a misconception, or (c) appeared to us to be
``Lispy'' and hence not of broad interest.

\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
list the final catalog.
For each, we also
present a Tutor question
for which the marked wrong answer can be explained by \emph{only} the
named misconception.
That is, that program-answer pair is a representative example of that
misconception.\footnote{An astute reader may well imagine another
  misinterpretation; but we presumably did not find evidence of it in our
  data. We do welcome other misinterpretations, for which we
  can add more misinterpreters!}

@@

The misinterpreters helped us find misinterpretations that we had
overlooked. For instance, consider this program from \quizState:
\begin{lstlisting}
(defvar x 12)
(deffun (f x)
  (set! x 0))
(f x)
x
\end{lstlisting}
We had assumed the wrong answer
\texttt{0} is caused by \miscon{CallByRef}.
However, our misinterpreters made us realize it can also be explained
by \miscon{FlatEnv}. This could also explain why we see a difference in
error rate when the formal and actual parameter have the same
name (as mentioned in \Cref{s:misconceptions-quiz:state}).

@@

The problem here is that the word ``redefine'' underspecifies how the
second definition is interpreted. We had interpreted it as
\textit{mutating} the binding established by the first definition,
which fits \miscon{DefOrSet}. However, another possibility is that
it \emph{shadows} the binding (i.e., establishes a new scope block). We
did not recognize that the original misconception is underspecified
until we uncovered the new (surmised) misconception.

@@

To summarize, we used the SMoL Tutor, with an expanded set of
programs, to investigate student misconceptions. We implemented the
idea of misinterpreters to help us properly classify student
performance. We were able to reconfirm all the previously identified
misconceptions, refine some of them, and also identify potential new
ones (that need further investigation).

@@

\end{description}
To study this, we perform the following analysis per
misconception. Using misinterpreters, we identify those questions
whose wrong answers fit only one misconception (i.e., only one output
matches that produced by that misinterpreter, and it matches only that
misinterpreter). We then examine student performance over time across
those questions. This would let us examine how they do on just that
topic, in isolation, over time.

We present the result of this analysis in two forms. Graphically, we
show plots in \Cref{f:tutor-trend-plots}. Each figure shows the percentage of students who
chose the answer corresponding to that misconception. Ideally, we
would like to see these percentages diminish.

@@

We also perform a logistic regression to see whether these
improvements are significant (at a $p < 0.05$ threshold); details are
in \supplRef{Paper.html}. Of the 11:
\begin{itemize}

@@

We therefore view the above data as
purely formative: they suggest that the Tutor \emph{most likely did
  not do harm} and perhaps even \emph{may have done some
  good}. However, it would be improper to read too much into the
analysis. It is quite possible that some of the other problems would
have found issues. Rather, what we really see is value in the
misinterpreter concept. It is not only useful for analysis, it is also
valuable for \emph{problem design}: in the next iteration of the
Tutor, we will use misinterpreters actively to shape the incorrect answers and,
as necessary, update the programs as well. We therefore hope to have
much more thorough analyses in future work.

@@

There is, of course, a significant danger that the data above have
been overfitted to only one institution (\Cref{s:univ-1-course},
henceforth University 1),
thereby actually reflecting the state of its curriculum rather than
some greater truth about program behavior understanding.  We already
have some reason to believe this is not the case: the related work
discussed in \Cref{s:rel-work} is drawn from many institutions in
multiple countries with different educational preparations, levels,
and demographics. Nevertheless, that gives us only
limited information about the specific questions and misconceptions
described above.

@@

There are some small differences.
\cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
that is different from \miscon{FlatEnv}:
\begin{description}
  \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
  When a function is called, the function body is evaluated in an
  environment that extends from the \textit{caller's} environment.
\end{description}
We don't include \surmisedMiscon{CallerEnv} in our analysis because in our
data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
are also explainable by \miscon{FlatEnv}.

@@

Our idea of ``goal sentences'' is not substantially different from the
\textit{rules of program behavior} from \cite{duranRulesProgramBehavior2021}.
We only used these sentences (or rules) to guide the
design of the Tutor and as text in the Tutor itself.
\cite{duranRulesProgramBehavior2021} argue for other uses of such
sentences.

@@

A particularly notable threat is the use of Lispy
syntax. We chose it for two reasons: both because it helps clarify
scope (\Cref{s:smol}) and because the rest of the course used
Racket. However, students may well perform differently with more
familiar syntaxes. It seems unlikely their performance would be \emph{too}
different given the many languages that have produced similar
misconceptions (\Cref{t:rel-work}). Nevertheless, we intend to use a
variety of syntaxes to examine this issue further.

@@

Is our reasoning valid? We have applied standard techniques and
measurements for evaluating student responses. However, our
instruments still lack the validity of a proper concept
inventory. Their creation requires heavyweight processes (such as
Delphi methods \cite{goldmanSettingScopeConcept2010}) that require many hours of expert attention as
well as conversations with learners, and can hence be prohibitive in
cost. The Quizius method (\Cref{s:quizius}) was created precisely to
be an inexpensive method that provides a good proxy.

>>>>

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}

@@

  In this proposal, I present a curated list of misconceptions, and an
  instrument to detect them. These are distilled from student work
  over several years and match and extend prior research.  I also
  present an automated, self-guided tutoring system. The tutor builds
  on strategies in the education literature and is explicitly designed
  around identifying and correcting misconceptions.

  I have tested the tutor in multiple settings.
  Our data consistently show that (a) the misconceptions I tackle are
  widespread, and (b) the tutor appears to improve understanding.

  [FILL future work]
\end{abstract}

@@

\begin{description}
  \item[RQ1] What are the misconceptions that are about program
  behavior, apply to most modern programming languages, and present
  (even) in students with prior programming backgrounds?
  \item[RQ2] What would be a tutoring system (the SMoL Tutor) that
  draws on cognitive and educational psychology concepts, utilizes a
  notional machine (the Stacker), and aims at correcting the
  misconceptions?
  \item[RQ3] How effective is SMoL Tutor at correcting the
  misconceptions?
\end{description}

@@

  \item I provide a collection of brief programs for detecting
        misconceptions about the features of SMoL and their
        interaction. These programs are short and most are readily
        translatable to a wide variety of programming languages that
        share the SMoL characteristics.

  \item I show that multiple populations have problems with these
        programs.

  \item I present a list of curated misconceptions generated
        after multiple rounds of analysis.

  \item I implement these misconceptions as interpreters, which
        find weaknesses in our manual analysis and hold potential for
        generative use in the future.

  \item I present a tutoring system, the SMoL Tutor, that measurably
        corrects these misconceptions.

@@

This document first presents the current results and ends with future
work \& timeline (\cref{s:future-work}).

\paragraph{A Note on Terminology} I use the term
``behavior'' to refer to the meaning of programs in terms of the
answers they produce. A more standard term for this would, of course, be
``semantics''. However, the term ``semantics tutor'' might
mislead some readers into thinking it
teaches people to read or write a formal semantics, e.g., an
introduction to ``Greek'' notation. Because that is not the kind of
tutor I are describing, to avoid confusion, I use the term
``behavior'' instead.

@@

There is an extensive body of literature on tutoring systems (\cite{vanlehnBehaviorTutoringSystems2006a} is
a quality survey), and indeed whole conferences are dedicated to
them. I draw on this literature. In particular, it is common in the
literature to talk about a ``two-loop'' architecture \cite{vanlehnBehaviorTutoringSystems2006a} where the outer
loop iterates through ``tasks'' (i.e., educational activities) and the inner loop iterates through UI events within a task. I follow the same
structure in our tutor (\Cref{s:tutor-uiux}).

Many tutoring systems focus on teaching programming
(such as the well-known and heavily studied LISP Tutor \cite{andersonLISPTutor1985}), and in
the process undoubtedly address some program behavior
misconceptions. Our SMoL Tutor differs in a notable way: it does not
try to teach programming per se. Instead, it assumes a basic
programming background and focuses entirely on program behavior misconceptions
and correcting them. I are not aware of a tutoring system (in computer science) that has
this specific design.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:univ-1-course}) using Racket
\cite{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
I have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

Nevertheless, most SMoL programs are easy to translate to other
languages. I present machine-translated Python and JavaScript
versions of our programs in \supplRef{Translated\_Programs.html}. Furthermore, we intend to
make a multi-lingual tutor in the future (\Cref{s:future-work}).

@@

Students are given a working implementation of SMoL, built
as a \verb|#lang| language inside Racket \cite{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

\section{Proposal Roadmap}
\label{s:roadmap}
\label{s:univ-1-course}

This proposal describes a four-year effort to obtain a quality instrument
to measure misconceptions and to build a tutor to address them.

@@

I then present the tutor (\Cref{s:system-tutor}). In particular, we
present the list of misconceptions---grounded in student data---that
are the focus of this project (\Cref{s:misconceptions-tutor}). I also
evaluate the tutor's effectiveness at correcting them
(\Cref{s:effect-tutor}).

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \cite{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \cite{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \verb|sqrt(4)|?'', then \verb|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \verb|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. I expect our
  instruments to be useful in the same way.}

All these, however, add up to a somewhat challenging demand. I want
to produce a list of questions (each one an MCQ) such that
\begin{enumerate}

@@

\end{itemize}
and so on.
I therefore manually curated the Quizius output
to address these issues.

@@

I therefore turned the programs into a set of quizzes (in the US
sense: namely, a brief test of knowledge) that we call the SMoL
Quizzes. There were three quizzes, ordered by linguistic
complexity. The first consisted of only basic operators and
first-order functions. The second added variable and structure
mutation. The third added \lstinline|lambda| and higher-order functions.

@@

Question orders were partially randomized. I wanted students to get
some easy, warm-up questions initially, so those were kept at the
beginning. Similarly, we wanted programs that are syntactically
similar to stay close to each other in the quiz. This is so that, when
students got a second such program, they would not have to look far to
find the first one and confirm that they are indeed (slightly)
different, rather than wonder if they were seeing the same program
again.

@@

Due to space limitations, we present the entire instrument in \supplRef{SMoL Quizzes/instrument}.
Here we focus on a few programs where student choices correspond to
misconceptions identified earlier, thereby also showing
that the curated programs are
still effective.
Syntactically, \verb|#t| and \verb|#f| are true and false,
while \verb|#(...)| is a vector. I use a * to indicate the correct
answer (which also matches the implementation's output).

@@

So far, we have focused on identifying problems. As noted earlier
(\Cref{s:smol-quizzes}), we provided students with refutation texts after the quizzes,
but it is unclear to what extent students read, understood, or
internalized these. I also wanted to determine whether other
populations run into these issues, and it was unclear whether we would
be able to administer passive quizzes to them.

Furthermore, while the quizzes may be a useful diagnostic, our goal is
not only to find faults but to improve the understanding of basic
programming language behavior. I believe (and presumably so does
anyone else who writes a formal semantics!)\ that an understanding of
these basic program behaviors is important, and even more so when it
comes to understanding concurrency, ownership, and other advanced
features. Even more simply, misunderstanding these clearly impacts
development and debugging time.

In response, we decided to create a \emph{tutor}: the SMoL Tutor. It is built
around our quiz instruments and the detected misconceptions. We
describe the Tutor from a user's perspective in \Cref{s:tutor-uiux}, and
explain how it was populated from the
SMoL Quizzes in \Cref{s:tutor-refine-program-set}.
I then discuss what
we learned from its data (\Cref{s:misconceptions-tutor}), and finally
evaluate its effectiveness as a \emph{tutor} (\Cref{s:effect-tutor}).

@@

In addition to asking students questions, the Tutor along the way
introduces terminology and states the true conceptions. These are the
teaching goals of the Tutor. I therefore refer to these as
\newterm{goal sentences}. \Cref{t:tutor-goal-sentences} lists
the (abbreviated) goal sentences for each tutorial.
Readers can find the full Tutor in \supplRef{SMoL Tutor/instrument}.

@@

The SMoL Tutor's interpreting questions are the final instrument of
this paper.
I bootstrap it from the SMoL
Quizzes, but made the following alterations:
\begin{enumerate}

@@

  \item I renamed vector operators to avoid confusion between
        \lstinline|vset!|, which replaces a vector element, and \lstinline|set!|,
        which mutates variables. I rename all vector operators from, for example,
        \lstinline|vset!| to \lstinline|vec-set!|.

  \item I deferred the local binding forms to the end, since we felt
        they were less essential. I therefore rewrote programs that use
        it to not depend on it.

        % \item I removed some questions that we deemed ``Lispy'' (e.g.,
        %   involving the difference between parenthetical and infix syntax).

  \item I removed some programs that relied on underspecified aspects or
        unimportant aspects of SMoL. For example, some depended on
        whether operations on pairs could be applied to arbitrary vectors:

@@

  \item I resolved ambiguities in some programs, either adding
    answer choices or even adding other questions to tease
    out different interpretations.

@@

  \item I removed questions related to function equality, which was
        not a focus of the Tutor.\footnote{However, we believe there is
          a good deal of confusion about notions of equality. We
          intend to
          add that as a major tutorial component in future versions.}

  \item I removed programs of a ``Lispy'' nature, such as one where
    the answer depended on whether the reader correctly understood
    this inequality:

@@

It was important to add wrong answers that are not utterly
implausible, because those would become easy to eliminate. Therefore,
we added numeric constants mentioned in the problem, permuted some of
the values in case of multiple outputs, and so on. In general, we
ended up increasing the number of choices substantially:
only 8 out of the 89 questions have three or four choices; 70
questions have 5--8 choices; and 11 questions have 9--14 choices. I hope
this reduced both guessing and elimination, and forced students to
actually think through the program. Of course, these new answers do
not have a clear associated misconception, so their mistakes are given
the generic explanation.

@@

I now examine what we learned from the interpreting
questions in terms of program behavior
misconceptions. But first, we explain how we associate incorrect
answers with misconceptions.

I do not present the results of \textbf{begin} and
\textbf{local} (\Cref{t:tutor-goal-sentences}), because they focus on constructs not usually found in
non-Lispy languages. They were included in the Tutor to help students
write programs for the course, but are not interesting from a SMoL
perspective.

@@

By running programs through the misinterpreters, we can more uniformly
and rigorously identify \emph{all} the misconceptions associated with
a wrong answer. Furthermore, if we identify a new misconception (or
alter a misinterpreter), it is easy to automatically re-classify all
the answers. By using misinterpreters, we indeed found new
interpretations for existing program-answer pairs.
I provide all the misinterpreters in the artifact.

@@

I iteratively created
our final catalog of misconceptions
(from the perspective of the data in this paper).
I started
with misinterpreters representing the misconceptions described in
\Cref{s:misconceptions-quiz}. These are misconceptions for which we
have reasonable validation (due to the prose in Quizius), so we call
them \emph{grounded} misconceptions. I then looked at wrong answers
not covered by these but chosen by students, and did our best to
distill these into misconceptions. These are \emph{surmised}
misconceptions (which we
identify with a $^\ddag$), which need to be validated in the future.\footnote{To keep
  the time spent in each module reasonable, the SMoL Tutor does not
  ask students for rationales for their programs. However, we could
  easily modify it to do so a small number of times per user, to make
  progress towards this need.} I then re-ran the misinterpreters
against the chosen answers. I terminated when all the remaining wrong
answers were either (a) found in very few students (we found a gap
between 23\% and 13\%, and hence took 14\% as the threshold),
(b) difficult for
us to attribute to a misconception, or (c) appeared to us to be
``Lispy'' and hence not of broad interest.

\Cref{t:tutor-grounded-misconceptions-1,t:tutor-grounded-misconceptions-2,t:tutor-surmised-misconceptions}
list the final catalog.
For each, we also
present a Tutor question
for which the marked wrong answer can be explained by \emph{only} the
named misconception.
That is, that program-answer pair is a representative example of that
misconception.\footnote{An astute reader may well imagine another
  misinterpretation; but we presumably did not find evidence of it in our
  data. I do welcome other misinterpretations, for which we
  can add more misinterpreters!}

@@

The misinterpreters helped us find misinterpretations that we had
overlooked. For instance, consider this program from \quizState:
\begin{lstlisting}
(defvar x 12)
(deffun (f x)
  (set! x 0))
(f x)
x
\end{lstlisting}
I had assumed the wrong answer
\texttt{0} is caused by \miscon{CallByRef}.
However, our misinterpreters made us realize it can also be explained
by \miscon{FlatEnv}. This could also explain why we see a difference in
error rate when the formal and actual parameter have the same
name (as mentioned in \Cref{s:misconceptions-quiz:state}).

@@

The problem here is that the word ``redefine'' underspecifies how the
second definition is interpreted. I had interpreted it as
\textit{mutating} the binding established by the first definition,
which fits \miscon{DefOrSet}. However, another possibility is that
it \emph{shadows} the binding (i.e., establishes a new scope block). We
did not recognize that the original misconception is underspecified
until we uncovered the new (surmised) misconception.

@@

To summarize, we used the SMoL Tutor, with an expanded set of
programs, to investigate student misconceptions. I implemented the
idea of misinterpreters to help us properly classify student
performance. I were able to reconfirm all the previously identified
misconceptions, refine some of them, and also identify potential new
ones (that need further investigation).

@@

\end{description}
To study this, we perform the following analysis per
misconception. Using misinterpreters, we identify those questions
whose wrong answers fit only one misconception (i.e., only one output
matches that produced by that misinterpreter, and it matches only that
misinterpreter). I then examine student performance over time across
those questions. This would let us examine how they do on just that
topic, in isolation, over time.

I present the result of this analysis in two forms. Graphically, we
show plots in \Cref{f:tutor-trend-plots}. Each figure shows the percentage of students who
chose the answer corresponding to that misconception. Ideally, we
would like to see these percentages diminish.

@@

I also perform a logistic regression to see whether these
improvements are significant (at a $p < 0.05$ threshold); details are
in \supplRef{Paper.html}. Of the 11:
\begin{itemize}

@@

I therefore view the above data as
purely formative: they suggest that the Tutor \emph{most likely did
  not do harm} and perhaps even \emph{may have done some
  good}. However, it would be improper to read too much into the
analysis. It is quite possible that some of the other problems would
have found issues. Rather, what we really see is value in the
misinterpreter concept. It is not only useful for analysis, it is also
valuable for \emph{problem design}: in the next iteration of the
Tutor, we will use misinterpreters actively to shape the incorrect answers and,
as necessary, update the programs as well. I therefore hope to have
much more thorough analyses in future work.

@@

There is, of course, a significant danger that the data above have
been overfitted to only one institution (\Cref{s:univ-1-course},
henceforth University 1),
thereby actually reflecting the state of its curriculum rather than
some greater truth about program behavior understanding.  I already
have some reason to believe this is not the case: the related work
discussed in \Cref{s:rel-work} is drawn from many institutions in
multiple countries with different educational preparations, levels,
and demographics. Nevertheless, that gives us only
limited information about the specific questions and misconceptions
described above.

@@

There are some small differences.
\cite{fleuryParameterPassingRules1991} identifies a dynamic scope misconception
that is different from \miscon{FlatEnv}:
\begin{description}
  \item[\surmisedMiscon{CallerEnv}] Function values don't remember their environments.
  When a function is called, the function body is evaluated in an
  environment that extends from the \textit{caller's} environment.
\end{description}
I don't include \surmisedMiscon{CallerEnv} in our analysis because in our
data, all wrong answers that can be explained by \surmisedMiscon{CallerEnv}
are also explainable by \miscon{FlatEnv}.

@@

Our idea of ``goal sentences'' is not substantially different from the
\textit{rules of program behavior} from \cite{duranRulesProgramBehavior2021}.
I only used these sentences (or rules) to guide the
design of the Tutor and as text in the Tutor itself.
\cite{duranRulesProgramBehavior2021} argue for other uses of such
sentences.

@@

A particularly notable threat is the use of Lispy
syntax. I chose it for two reasons: both because it helps clarify
scope (\Cref{s:smol}) and because the rest of the course used
Racket. However, students may well perform differently with more
familiar syntaxes. It seems unlikely their performance would be \emph{too}
different given the many languages that have produced similar
misconceptions (\Cref{t:rel-work}). Nevertheless, we intend to use a
variety of syntaxes to examine this issue further.

@@

Is our reasoning valid? I have applied standard techniques and
measurements for evaluating student responses. However, our
instruments still lack the validity of a proper concept
inventory. Their creation requires heavyweight processes (such as
Delphi methods \cite{goldmanSettingScopeConcept2010}) that require many hours of expert attention as
well as conversations with learners, and can hence be prohibitive in
cost. The Quizius method (\Cref{s:quizius}) was created precisely to
be an inexpensive method that provides a good proxy.

Ultimately, our goal is to provide a reasonable instrument for
widespread use. While the set of all misconceptions could be
unbounded, we believe our tasks, especially as embedded in the Tutor,
provide a good starting point for others. In particular, if students
select a wrong answer, that is still of \emph{some} use to an
educator, even if the precise misconception cannot be pinned down with
the highest accuracy. I therefore believe our instruments, and our
misinterpreter technique, are of general value.
